---
title: "GLM for binary data"
author: "Laure"
date: "8 jan 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Binary Data

## Binary data 

Binary data may occur in two forms 

- ungrouped in which the variable can take one of two values, say success/failure 

- grouped in which the variable is the number of successes in a given number of trials 

The natural distribution for such data is the Binomial(n,p) distribution, where in the ﬁrst case n = 1

## Models for binary data 

Suppose $Yi ∼ Binomial(n_i,p_i)$ and we wish to model the proportions $Y_i/n_i$. Then
$$E(Y_i/n_i) = p_i$$ 
$$ var(Y_i/n_i) = \frac{1}{n_i} p_i(1 − p_i) $$
So our variance function is $V(µ_i) = µ_i(1−µ_i)$ Our link function must map from (0,1)→(−∞,∞). A common choice is $g(µi) = logit(µi) = log(\frac{µi}{1−µi})$. 

Binomial data can be modelled by a GLM, with the canonical logit link. This model is known as the logistic regression model and is the most popular for binary data. There are two other links commonly used in practice 

- probit link $g(µ_i) = Φ^{−1}(µ_i)$ where $Φ$ denotes the cumulative distribution function of N(0,1) 

- complementary log-log link $g(µ_i) = log(−log(1−µ_i))$

The logit and probit functions are symmetric and - once their variances are equated - are very similar. Therefore it is usually diﬃcult to choose between them on the grounds of ﬁt. The logit is usually preferred over the probit because of its simple interpretation as the logarithm of the odds of success (pi/(1−pi)). The complementary log-log is asymmetric and may therefore be useful when the logit and probit links are inappropriate. We will concentrate on using the logit link.

```{r, echo = T,warning= T}
# Comparing the different link functions 
mu.logit <- function(eta) 1/(1 + exp(-eta)) 
mu.probit <- function(eta) pnorm(eta, 0, pi/sqrt(3)) 
mu.cloglog <- function(eta) 1 - exp(-exp(eta))

plot(mu.logit, (-4): 4, xlim = c(-4, 4), ylim = c(0,1), xlab = expression(eta), ylab = expression(mu == g^-1 * (eta))) 
curve(mu.probit, (-4):4, add = TRUE, lty = 2) 
curve(mu.cloglog, (-4):4, add = TRUE, lty = 3) 
legend(-4, 1, c("logit", "probit", "complementary log-log"), lty = 1:3)
```

## Interpretation 

Consider the logistic regression model 
$$log(\frac{p_i}{1 − p_i})= β_0 + β_1 x_{1i}$$
If we increase x1 by one unit 
$$ log( \frac{pi}{1−pi}) = β_0 + β_1(x_{1i} + 1) = β_0 + β_1 x_{1i} + β_1$$
So 
$$\frac{p_i}{1 − p_i} = exp(β_0 + β_1 x_{1i})exp(β_1)$$
the odds are multiplied by $exp(β_1)$.

# Implementation in R

## Case control study data 

In the mid-80s a case-control study on risk factors for malignant melanoma was conducted in Denmark (Osterlind et al. The Danish case-control study of cutaneous malignant melanoma I: Importance of host factors. Int J Cancer 1988; 42: 200-206).
The cases were patients with skin melanoma (excluding lentigo melanoma), newly diagnosed from 1 Oct, 1982 to 31 March, 1985, aged 20-79, from East Denmark, and they were identified from the Danish Cancer Registry.
The controls (twice as many as cases) were drawn from the residents of East Denmark in April, 1984, as a random sample stratified by sex and age (within the same 5 year age group) to reflect the sex and age distribution of the cases.

| Variable            | Units or Coding                                    | Type       | Name    |
|:--------------------|:---------------------------------------------------|:-----------|:--------|
| Case-control status | 1=case, 0=control                                  | numeric    | cc      |
| Sex                 | 1=male, 2=female                                   | numeric    | sex     |
| Age at interview    | age in years                                       | numeric    | age     |
| Skin complexion     | 0=dark, 1=medium, 2=light                          | numeric    | skin    |
| Hair colour         | 0=dark brown/black, 1=light brown, 2=blonde, 3=red | numeric    | hair    |
| eye colour          | 0=brown, 1=grey, green, 2=blue                     | numeric    | eyes    |
| Freckles            | 1=many, 2=some, 3=none                             | numeric    |freckles |
| Naevi, small        | no. naevi < 5mm                                    | numeric    | nvsmall |
| Naevi, largs        | no. naevi >= 5mm                                   | numeric    | nvlarge |

```{r, echo = F,warning= T}
## load packages 
library(Epi)
library(dplyr)
```

```{r, echo = T,warning= T}
# reading Data 
mel <- read.table("http://bendixcarstensen.com/SPE/data/melanoma.dat", header = T, na.string = ".")
str(mel)
head(mel)
```

```{r, echo = T,warning= T}
# Housekeeping 
# Convert sex, skin, hair, eye into factor variables
source("http://bendixcarstensen.com/SPE/data/melanoma-house.r")
str(mel)
```
## Exploring binary data 

```{r, echo = T,warning= T}
# Exploring binary data 
# When the explanatory variables are categorical, a simple approach is to calculate proportions within subgroups of the data. 
# When some of the explantory variables are continuous, plots can be more helpful. 

# skin
with(mel, table(cc,skin)) # number of cases and controls by skin color
stat.table(index = skin, contents = ratio(cc,1-cc), data = mel) # odds of being a case by skin color
# eyes
with(mel, table(cc,eyes))
stat.table(index = eyes, contents = ratio(cc,1-cc), data = mel)
# freckles
with(mel, table(cc,freckles))
stat.table(index = freckles, contents = ratio(cc,1-cc), data = mel)
# age 
# plot(cc ∼ age, data = mel) # Simple scatterplot are not that helpful 
plot(jitter(cc, 0.1) ∼ age, data = mel) # can be improved by using jitter
boxplot(age ∼ cc, xlab = "Case/Control (1/0)", ylab = "Age", data = mel) # an alternative to scatter plot is to use a box plot 

# Note crude estimates of the odds ratio can be derived for the categorical data from tables above
```

## Modeling binary data 

```{r, echo = T,warning= T}
# Calculate effects of an exposure on a response

# Effects are calculated as differences in means for a metric response,
# odds ratios/relative risks for a binary response, 
# and rate ratios/rate differences for a failure or count response.

# The k-1 effects for a categorical exposure with k levels are relative to a baseline which, by default, is the first level.

effx(response = cc, type ="binary", exposure = skin, data = mel)
effx(response = cc, type ="binary", exposure = eyes, data = mel)
effx(response = cc, type ="binary", exposure = freckles, data = mel)

# The function effx gives the same ouptput as using GLM with binomial family and logit link
```

```{r, echo = T,warning= T}
# Modeling for binary data 

# GLM with binomial family and logit link
mf <- glm(cc ~ freckles, family ="binomial", data = mel) 
# Summary
summary(mf)
# Odds ratio/relative risks  
round(ci.exp( mf ),2)

# Controlling with age and sex 
# Using effx
effx(cc, type = "binary", exposure = freckles, control = list(age.cat,sex), data = mel)
# Using glm
mfas <- glm(cc ~ freckles + age.cat + sex, family = "binomial", data = mel) # this gives same result as using effx with control
round(ci.exp(mfas),2)
summary(mfas)
```


```{r, echo = T,warning= T}
# Testing for no effect of an exposure 
# to test for no effect at all of freckles you need a likelihood ratio test.
# This involves fitting two models, one without freckles and one with, and recording the change in deviance.

# Because of NAs in freckles we need to restrict the 1st model to people who have non NA freckles
mas <- glm(cc ~ age.cat + sex, family = "binomial", data = subset(mel, !is.na(freckles)) )
anova(mas, mfas, test = "Chisq")
```

The change in residual deviance is 1785.9 - 1737.1 = 48.8 on 1389 -􀀀 1387 = 2 degrees of
freedom. The P-value corresponding to this change is obtained from the upper tail of the
cumulative distribution of the 2-distribution with 2 df:

```{r, echo = T,warning= T}
1 - pchisq(48.786, 2)
```


